Technical Report for **Thermal reactionomes reveal divergent responses to thermal extremes in warm and cool-climate ant species**

### 2015-09-30 ApTranscriptome_TR.Rmd
added resampling analysis: greatly reduces number of significant transcripts

---
output: pdf_document
---

Technical Report for **Thermal reactionomes reveal divergent responses to thermal extremes in warm and cool-climate ant species**
========================================================================================

**Author:** [John Stanton-Geddes](john.stantongeddes.research@gmail.com)

**`r format(Sys.time(), '%d %B, %Y')`**

**Technical Report No. 3**

**Department of Biology**

**University of Vermont**

@@

# Load Bioconductor libraries
#source("http://bioconductor.org/biocLite.R")
#biocLite("topGO")
#biocLite("Rgraphviz")
#biocLite("GOSemSim")
library(topGO) 
library(Rgraphviz)	
library(GOSemSim)

@@

We quantify the relative importance of each of these mechanisms to thermal adaptation in these two species by evalutating differences in the reactionomes between species. We use gene set enrichment analysis to provide additional insight on the cellular processes underlying the transcriptome-wide changes in expression.

@@

Note that I used the reduced assembly, prior to cleaning out contaminants. As I use the "clean" assembly for read mapping and identification of responsive genes, the contaminants are dis-regarded in downstream analyses.

@@

Note that expression levels at each temperature treatment are highly correlated between the two colonies.

@@

With gene expression quantified using all reads, I now remove the transcripts identified as contaminants by DeconSeq.

@@

# read file
cont.list <- read.table("results/trinity-full/cont.list")
str(cont.list)

@@

Many of these are likely false positives, so I adjust P-values using false discovery rate (FDR). Only those transcripts with less than 5% FDR are retained as significant.

@@

# subset to significant transcripts
signif.transcripts <- RxNpval[which(RxNpval$padj < 0.05), ]

@@

At the 5% FDR significance threshold, there are `r nrow(signif.transcripts)` transcripts with an overall significant model.

(2) Fit linear model to overall significant transcripts; perform stepAIC to retain only significant terms, and save `lm` output to list

@@

```{r expression_shape, eval=TRUE, echo=TRUE}
# calculate response type for responsive transcripts
interaction.response.type <- ldply(interaction.lms, .progress = "none", .inform = TRUE, RxNtype)
stopifnot(nrow(interaction.response.type) == length(interaction.lms))

# calculate response types for transcripts without interactions
temperature.response.type <- ldply(temperature.lms, .progress = "none", .inform = TRUE, RxNtype)
stopifnot(nrow(temperature.response.type) == length(temperature.lms))

# for subsequent analyses, focus on the transcripts with significant temperature x species interaction only
colnames(interaction.response.type)[which(colnames(interaction.response.type) == ".id")] <- "Transcript"
interaction.response.type <- data.table(interaction.response.type)
setkey(interaction.response.type, Transcript)
str(interaction.response.type)

# save results to file
write.table(file = paste(resultsdir, "Ap_responsive_transcripts_", Sys.Date(), ".txt", sep = ""), interaction.response.type, row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
```

@@

```{r temperature_response}
A22.type.table <- table(interaction.response.type$A22.type)
Ar.type.table <- table(interaction.response.type$Ar.type)

@@

```{r mosaic_plot}
# Reorganize for plotting to show overlap among categories between colonies
type.table <- table(Acar = interaction.response.type$Ar.type, Apic = interaction.response.type$A22.type)
# reorder
tt2 <- type.table[c("Low", "Intermediate", "High", "Bimodal", "NotResp"), c("Low", "Intermediate", "High", "Bimodal", "NotResp")]

# plot
png("mosaic_plot.png")
mosaic.colors <- brewer.pal(5, "Blues")
par(mar = c(5.1, 3.1, 2.1, 7.5))
mosaicplot(tt2, ylab = expression(italic("A. picea")), xlab = "", main = "", color = mosaic.colors, cex.axis = 0.01)
labels <- c("Low", "Intermediate", "High", "Bimodal", "NotResp")
text(c(.1,.5,.75,.85,.99), par("usr")[3] - 0.05, srt = 45, adj = 1, labels = labels, xpd = TRUE, font = 1)
text(.5, par("usr")[4], "A. carolinensis", xpd = TRUE, font = 3)
legend(1.05, .75, fill = mosaic.colors, c("Low", "Intermediate", "High", "Bimodal", "NotResp"), xpd=TRUE)
dev.off()
```

@@

```{r mh_test}
mh.test <- mh_test(as.table(tt2))

@@

# e2 (Acar Intermediate) contributes ~44% to Z~0~
```

@@

mh_plot <- qplot(x=Apic, y=Acar, data=md, fill=value, geom="tile", ylim = rev(levels(md$Acar))) +
  scale_fill_gradient2(limits=c(-700, 600), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(x = "A. picea", y = "A. carolinensis")

@@

mh_plot_sub1 <- qplot(y=Apic, x=Acar, data=md_sub1, fill=value, geom="tile", xlim = rev(levels(md_sub1$Acar))) + 
  scale_fill_gradient2(limits=c(-700, 600), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(y = "A. picea", x = "A. carolinensis") +
  theme(legend.position = "right") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

@@

mh_plot_sub2 <- qplot(y=Acar, x=Apic, data=md_sub2, fill=value, geom="tile", xlim = rev(levels(md_sub2$Apic))) + 
  scale_fill_gradient2(limits=c(-700, 600), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(y = "A. carolinensis", x = "A. picea") + 
  theme(legend.position = "none") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 2.5, ymax = 3.5, fill = "transparent", linetype=2, colour="black")

@@

```{r er_cartoon}
enhanced_response_dat <- md_sub1
enhanced_response_dat$value <- 0
# set non-shared response to random negative value 
enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Low"), "value"] <- runif(5, min=140, max=150)
enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Bimodal"), "value"] <- rep(-200, 5)
# change shared response to positive
enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Low" & enhanced_defense_dat$Apic == "Low"), "value"] <- 50
enhanced_defense_dat[which(enhanced_defense_dat$Acar == "Bimodal" & enhanced_defense_dat$Apic == "Bimodal"), "value"] <- 50

@@

# plot
enhanced_response_plot <- qplot(y=Apic, x=Acar, data=enhanced_response_dat, fill=value, geom="tile", xlim = rev(levels(enhanced_response_dat$Acar))) + 
  scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(y = "A. picea", x = "A. carolinensis") +
  theme(legend.position = "right") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

enhanced_response_plot <- ggplot(enhanced_response_dat, aes(y = Apic, x = Acar, fill=value)) + 
  geom_tile(colour = "#ffffff", size =1) +
  xlim(rev(levels(enhanced_response_dat$Acar))) + 
  scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(y = "A. picea", x = "A. carolinensis") +
  theme(legend.position = "right") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

@@

```{r tol_cartoon}
tolerance_dat <- md_sub1
tolerance_dat$value <- 0
# set non-shared response to random positive value 
tolerance_dat[which(tolerance_dat$Acar == "Low"), "value"] <- runif(5, min=140, max=160)
tolerance_dat[which(tolerance_dat$Acar == "Bimodal"), "value"] <- runif(5, min=140, max=150)
# change shared response to near zero
tolerance_dat[which(tolerance_dat$Acar == "Low" & tolerance_dat$Apic == "Low"), "value"] <- 20
tolerance_dat[which(tolerance_dat$Acar == "Bimodal" & tolerance_dat$Apic == "Bimodal"), "value"] <- 20

@@

# plot
tolerance_plot <- ggplot(tolerance_dat, aes(y=Apic, x=Acar, fill=value)) + 
  geom_tile(colour = "#ffffff", size =1) +
  xlim(rev(levels(enhanced_response_dat$Acar))) + 
  scale_fill_gradient2(limits=c(-500, 500), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(y = "A. picea", x = "A. carolinensis") +
  theme(legend.position = "none") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

@@

The number of thermally-responsive in each response category differed between the colonies, with the msot transcripts expressed at *Low* temperatures in both colonies. For *ApVT*, an equal number of transcripts are expressed at *High* and *Bimodal*, followed by *Intermediate* transcripts. For *AcNC*, transcripts expressed at *Intermediate* temperatures are next most common, followed by *High* and *Bimodal*.

Interestingly, nearly half of the *High* genes in *AcNC* are *Low* in *ApVT*, and vice versa. In contrast, most of the *Low* genes in one species are also *Low* in the other species.

@@

Table 4 shows the number of transcripts that fall into each expression type for each each species. The totals for each species include the `r length(temperature.lms)` transcripts that have consistent temperature responses between the two colonies.

An interesting observation from the matched observations plot (Fig. 1) is that for **NotResp** genes in *A. carolinensis* the **High** and **Bimodal** categories in *A. picea* are over-represented. Finding that the mean expression level of these genes in *A. carolinensis* is greater than the expression level at the rearing temperature (25°C) in *A. picea* would be consistent with genetic assimilation.

```{r gen_assimilation_high}
Ap.high.Ac.nr <- interaction.response.type[which(interaction.response.type$A22.type == "High" & interaction.response.type$Ar.type == "NotResp"), ]

# get mean expression level for Ar transcripts
Ap.high.Ac.nr.TPM <- TPM.dt.sub[Ap.high.Ac.nr$Transcript]
Ap.high.Ac.nr.mean.exp.NR <- ddply(Ap.high.Ac.nr.TPM, .(Transcript), summarize, meanTPM = mean(TPM))

# get expression at optimum for A22 transcripts
Ap.high.Ac.nr.opt.exp <- Ap.high.Ac.nr$A22.opt

t.test(Ap.high.Ac.nr.mean.exp.NR$meanTPM, Ap.high.Ac.nr.opt.exp)
```

Contrary to expectations, mean expression of the non-responsive *A. carolinensis* genes is less than expression of the **High** genes in *A. picea*.

Next, I checked the same for the **Bimodal** genes in *A. picea*.

```{r gen_assimilation_bim}
Ap.bim.Ac.nr <- interaction.response.type[which(interaction.response.type$A22.type == "Bimodal" & interaction.response.type$Ar.type == "NotResp"), ]

@@

As with the **High** transcripts, expression was much higher for the responsive *A. picea* transcripts than for the non-responsive *A. carolinensis* transcripts.

@@

In this section, I perform a number of comparisons of the thermal reactionomes between the *Ar* and *A22* species. Specifically, I:

- compare profiles of the temperature of maximum expression for thermally-responsive transcripts
- compare basal expression at optimal temperature for thermally-responsive genes
- compare degree of upregulation to constitutive expression
- compare the critical temperature of transcript upregulation, T~on~, between species for transcripts induced in response to high or low temperatures 
- compare the critical temperature of transcript downregulation, T~off~, between species for transcripts induced in response to high or low temperatures

Predict expression for responsive transcripts.

@@

Probability density function of peak expression for transcripts that differ in expression between *A22* and *Ar*

@@

The `t.test` fails to account for the many orders of magnitude difference in expression among transcripts, e.g. non-equal variances. This problem is the key issue in the analysis of differential expression `r citep(c("10.1186/1471-2105-11-94", "10.1038/nprot.2013.099"))`. As my goal is simply to determine if expression is typically greater at optimal temperatures (19.5 C) in *Ar* than *A22* for genes that are up-regulated at high temperatures in *A22*, I use a non-parametric Wilcoxon signed rank-test

@@

Consistent with expectations, there is greater expression at 25C for *Ar* than *A22* transcripts for the set of transcripts that are transcripts that are up-regulated at high temperatures in *A22*. Note that A22 had the larger library size so if this was due to TPM not correctly accounting for differences in reads between samples, we would expect to see a positive instead of negative value here.

@@

I extract the **High** transcripts for *A. picea* and examine the log ratio of inducibility against the log ratio of constitutive expression in *A. picea* over *A. carolinensis**. According to the *genetic assimilation hypothesis* there should be a negative relationship between these ratios. That is, transcripts with greater constitutive expression in *A. carolinensis* should have greater inducibility in response to **High** temperatures in *A. picea*.

@@

As with the previous comparison, I also find no evidence for the genetic assimilation hypothesis.

@@

Thermally-responsive genes could also differ in the critical temperature of gene induction. To examine this, I determine the temperature at which each responsive gene has the greatest increase or decrease in expression.

@@

For next analyses, extract list of gene names by response type.

@@

Calculate T~on~ for *High* genes in each species. I use the observed (rather than predicted) values for each species as by nature of fitting a quadratic function, the maximum change tends to be at the extremes (38.5 or 0) for the predicted values.

@@

Genes with increased expression at *High* temperatures are on average turned on 1°C higher in *AcNC* than *ApVT*.

Repeat analysis for *Low* genes.

@@

As with genes turned on at high temperatures, *A. carolinensis* **Intermediate** genes are down-regulated on at higher temperatures than *A. picea* **Intermediate** genes. Towards low temperatures, the mean temperature of downregulation was marginally higher for *A. carolinensis* than *A. picea*.

@@

Make a plot with all 4 gene activation comparisons.

@@

Does the mean expression level for all responsive genes differ among colonies?

@@

@@

```{r ann_responsive, echo=TRUE}
responsive.lms.ann.type <- signif.transcripts[names(responsive.lms)]
# combine responsive.lms.ann with "type" information
responsive.lms.ann.type <- merge(responsive.lms.ann.type, interaction.response.type[,c("Transcript", "A22.type", "Ar.type")], by = "Transcript", all = TRUE)
str(responsive.lms.ann.type)
write.csv(responsive.lms.ann.type, file = paste(resultsdir, "Ap_responsive_genes_", Sys.Date(), ".csv", sep = ""), row.names = FALSE)
```

@@

There are also `r nrow(hsp_responsive)` heat shock related genes in the responsive transcripts, out of `r nrow(hsp_all)` total heat shock related genes in the transcriptome annotation.

@@

I use [topGO](http://www.bioconductor.org/packages/2.12/bioc/html/topGO.html) to perform gene set enrichment analysis (GSEA) seperately for each expression type (bimodal, intermediate, high, low).

First need to create gene ID to GO term map file

@@

then read map file.

@@

Using this gene2GO map file, perform GSEA for:

@@

Use `selectFDR` function to select transcripts with adjusted P < 0.05.

Perform GSEA for all thermally-responsive transcripts.

For many of the responsive categories, it appears that there is considerable redundancy in the enriched GO terms. I use the [GOSemSim](http://www.bioconductor.org/packages/release/bioc/html/GOSemSim.html) package to determine semantic similarity of enriched GO terms `r citep("10.1093/bioinformatics/btq064")` and then perform hierarhichal clustering based on this information distance.

@@

As the molecular processes involved in cold and hot temperature tolerance are different, I perform GSEA separately for each response type and compile results in a single table.

```{r gsea_Ap_high, results = 'hide'}
## Genes with *High* expression in both colonies
Ap.high <- interaction.response.type[which(interaction.response.type$Ar.type == "High" & interaction.response.type$A22.type == "High"), "Transcript"]
Ap.geneList.high <- rep(0, length = length(Ap.geneList))
names(Ap.geneList.high) <- names(Ap.geneList)
Ap.geneList.high[(which(names(Ap.geneList.high) %in% Ap.high))] <- 1
# check correct number of values set to 1
table(Ap.geneList.high)
# run GSEA
Ap.high.gsea <- gsea(genelist = Ap.geneList.high, geneID2GO = geneID2GO, plotpath = NA)
```

```{r gsea_Ap_high_cluster}
# similarity among enriched GO terms
Ap.high.GO.term.sim <- termSim(Ap.high.gsea$GO.ID, Ap.high.gsea$GO.ID, ont = "BP", organism = 'fly')
# replace GO IDs with terms
rownames(Ap.high.GO.term.sim) <- Ap.high.gsea$Term
colnames(Ap.high.GO.term.sim) <- Ap.high.gsea$Term

# distance matrix
Ap.high.hclust <- hclust(dist(Ap.high.GO.term.sim))
plot(Ap.high.hclust)

# report items in each cluster
Ap.high.hclust.report <- GSEAReportClusters(Ap.high.hclust, h = 1)

# add cluster information
Ap.high.gsea$Cluster <- NA

for(i in 1:length(Ap.high.hclust.report)) {
    Ap.high.gsea[Ap.high.hclust.report[[i]], "Cluster"] <- i
}
```

>>>>

---
title: Technical Report for **Thermal reactionomes reveal divergent responses to thermal extremes in warm and cool-climate ant species**
author: [John Stanton-Geddes](john.stantongeddes.research@gmail.com)
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document:
    keep_md: yes
---

@@

# Load Bioconductor libraries
#source("http://bioconductor.org/biocLite.R")
#biocLite("topGO")
library(topGO)

@@

We quantified the relative importance of each of these mechanisms to thermal adaptation in these two species by evalutating differences in the reactionomes between species. We used gene set enrichment analysis to provide additional insight on the cellular processes underlying the transcriptome-wide changes in expression.

@@

@@

Note that expression levels at each temperature treatment are highly correlated between the two colonies, indicating common patterns of expression response to temperature.

@@

With gene expression quantified using all reads, I removed the transcripts identified as contaminants by DeconSeq.

@@

# read file
cont.list <- read.table("results/trinity-full/cont.list")
# remote ">" leader
cont.list$V1 <- gsub(">", "", cont.list$V1)
str(cont.list)

@@

Many of these are likely false positives, so I adjusted P-values using false discovery rate (FDR).

@@

# subset to significant transcripts
pre.signif.transcripts <- RxNpval[which(RxNpval$padj < 0.05), ]
```

At the 5% FDR significance threshold, there are `r nrow(pre.signif.transcripts)` transcripts with an overall significant model.

Feedback from reviewers questioned whether this was sufficient to remove all false positives. To examine this, we performed a resampling simulation to determine how many transcripts would be identified as significant using the above approach. Specifically, we randomly re-assigned the expression levels to the temperature treatments for each species, fit the same linear model as for the real data, calculated the number of significant transcripts under the 5% FDR threshold, and repeated this 100 times. We then took the 95th percentile of the distribution of significant transcripts from the resampled datasets as the true null level of false positives.

As this resampling is slow, I ran it in a separate file (ApTranscriptome_resample.Rmd) and load the results here.

```{r resample}
# load resampling results
load("data/resample.Rda")
```

Using this approach, the empirical false positive, *after* adjusting *P*-values for FDR, is `r quantile (D$num_signif, probs = 0.95)`, indicating that only `r nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)` are true positives.

```{r}
# order by padj
pre.signif.transcripts <- pre.signif.transcripts[order(pre.signif.transcripts$padj), ]
# number to keep
keep <- nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)
signif.transcripts <- pre.signif.transcripts[1:keep, ]

@@

(2) Fit linear model to significantly responsive transcripts, perform stepAIC to retain only significant terms, and save `lm` output to list

@@

```{r expression_shape, eval=TRUE, echo=TRUE}
# calculate response type for all temperature responsive transcripts
sig.response.type <- ldply(responsive.lms, .progress = "none", .inform = TRUE, RxNtype)
colnames(sig.response.type)[which(colnames(sig.response.type) == ".id")] <- "Transcript"
sig.response.type <- data.table(sig.response.type)
setkey(sig.response.type, Transcript)
str(sig.response.type)

# calculate response type for transcripts with species x temperature interaction only
interaction.response.type <- ldply(interaction.lms, .progress = "none", .inform = TRUE, RxNtype)
colnames(interaction.response.type)[which(colnames(interaction.response.type) == ".id")] <- "Transcript"
interaction.response.type <- data.table(interaction.response.type)
setkey(interaction.response.type, Transcript)
str(interaction.response.type)

# calculate response types for transcripts without interactions
temperature.response.type <- ldply(temperature.lms, .progress = "none", .inform = TRUE, RxNtype)

# save results to file
write.table(file = paste(resultsdir, "Ap_responsive_transcripts_", Sys.Date(), ".txt", sep = ""), sig.response.type, row.names = FALSE, col.names = TRUE, quote = FALSE, sep = "\t")
```

@@

```{r temperature_response}
A22.type.table <- table(sig.response.type$A22.type)
Ar.type.table <- table(sig.response.type$Ar.type)

@@

## Test of marginal frequencies between transcripts in each expression category between species

@@

```{r mh_test}
type.table <- table(Acar = sig.response.type$Ar.type, Apic = sig.response.type$A22.type)
# reorder
tt2 <- type.table[c("Low", "Intermediate", "High", "Bimodal", "NotResp"), c("Low", "Intermediate", "High", "Bimodal", "NotResp")]

# McNemar's test
mh.test <- mh_test(as.table(tt2))

@@

# e2 (Acar Intermediate) contributes ~48% to Z~0~
```

@@

mh_plot <- qplot(x=Apic, y=Acar, data=md, fill=value, geom="tile", ylim = rev(levels(md$Acar))) +
  scale_fill_gradient2(limits=c(-200, 200), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(x = "A. picea", y = "A. carolinensis")

@@

mh_plot_sub1 <- qplot(y=Apic, x=Acar, data=md_sub1, fill=value, geom="tile", xlim = rev(levels(md_sub1$Acar))) + 
  scale_fill_gradient2(limits=c(-200, 200), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(y = "A. picea", x = "A. carolinensis") +
  theme(legend.position = "right") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

@@

mh_plot_sub2 <- qplot(y=Acar, x=Apic, data=md_sub2, fill=value, geom="tile", xlim = rev(levels(md_sub2$Apic))) + 
  scale_fill_gradient2(limits=c(-200, 200), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(y = "A. carolinensis", x = "A. picea") + 
  theme(legend.position = "none") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 2.5, ymax = 3.5, fill = "transparent", linetype=2, colour="black")

@@

```{r er_cartoon}
enhanced_response_dat <- md_sub1
enhanced_response_dat$value <- 0
# set non-shared response to random negative value 
enhanced_response_dat[which(enhanced_response_dat$Acar == "Low"), "value"] <- runif(5, min=-20, max=-10)
enhanced_response_dat[which(enhanced_response_dat$Acar == "Bimodal"), "value"] <- runif(5, min=-20, max=-10)
# change shared response to positive
enhanced_response_dat[which(enhanced_response_dat$Acar == "Low" & enhanced_response_dat$Apic == "Low"), "value"] <- 30
enhanced_response_dat[which(enhanced_response_dat$Acar == "Bimodal" & enhanced_response_dat$Apic == "Bimodal"), "value"] <- 30

@@

# plot
enhanced_response_plot <- ggplot(enhanced_response_dat, aes(y = Apic, x = Acar, fill=value)) + 
  geom_tile(colour = "#ffffff", size =1) +
  xlim(rev(levels(enhanced_response_dat$Acar))) + 
  scale_fill_gradient2(limits=c(-100, 100), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(y = "A. picea", x = "A. carolinensis") +
  theme(legend.position = "right") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

@@

```{r tol_cartoon}
tolerance_dat <- md_sub1
tolerance_dat$value <- 0
# set non-shared response to random positive value 
tolerance_dat[which(tolerance_dat$Acar == "Low"), "value"] <- runif(5, min=40, max=60)
tolerance_dat[which(tolerance_dat$Acar == "Bimodal"), "value"] <- runif(5, min=40, max=60)
# change shared response to near zero
tolerance_dat[which(tolerance_dat$Acar == "Low" & tolerance_dat$Apic == "Low"), "value"] <- -20
tolerance_dat[which(tolerance_dat$Acar == "Bimodal" & tolerance_dat$Apic == "Bimodal"), "value"] <- -20

@@

# plot
tolerance_plot <- ggplot(tolerance_dat, aes(y=Apic, x=Acar, fill=value)) + 
  geom_tile(colour = "#ffffff", size =1) +
  xlim(rev(levels(enhanced_response_dat$Acar))) + 
  scale_fill_gradient2(limits=c(-100, 100), low="#f1a340", high="#998ec3") +
  theme(axis.title = element_text(face="italic")) +
  labs(y = "A. picea", x = "A. carolinensis") +
  theme(legend.position = "none") + 
  geom_rect(xmin = 0.5, xmax = 1.5, ymin = 3.5, ymax = 4.5, fill = "transparent", linetype=2, colour="black") +
  geom_rect(xmin = 1.5, xmax = 2.5, ymin = 0.5, ymax = 1.5, fill = "transparent", linetype=2, colour="black")

@@

The number of thermally-responsive in each response category differed between the colonies. For *A. picea*, the majority of the responsive transcripts were in the *Low* category, followed by *Bimodal*, *High*, *Intermediate* and *Not Responsive*. For *A. carolinensis*, the majority of responsive transcripts were in the *Bimodal* and *High* categories, followed by *Intermediate*, *Low* and *Not Responsive*.

Interestingly, over half of the *High* genes in *AcNC* are *Low* in *ApVT*, and vice versa. In contrast, most of the *Low* genes in one species are also *Low* in the other species.

@@

Table 4 shows the number of transcripts that fall into each expression type for each each species. The totals for each species *do not* include the `r length(temperature.lms)` transcripts that have consistent temperature responses between the two colonies.

An interesting observation from the matched observations plot (Fig. 1) is that for **NotResp** genes in *A. carolinensis* the **Bimodal** category in *A. picea* is over-represented. Finding that the mean expression level of these genes in *A. carolinensis* is greater than the expression level at the rearing temperature (25°C) in *A. picea* would be consistent with genetic assimilation. That is, they have evolved to be constitutively activated in *A. carolinensis* and thus are no longer thermally-responsive as they are in *A. picea*.

```{r gen_assimilation_bim}
Ap.bim.Ac.nr <- sig.response.type[which(sig.response.type$A22.type == "Bimodal" & sig.response.type$Ar.type == "NotResp"), ]

@@

This hypothesis was not supported as expression was higher for the **Bimodal** transcripts in *A. picea* than the non-responsive *A. carolinensis* transcripts.

@@

In this section, I performed a number of comparisons of the thermal reactionomes between the *Ar* and *A22* species. Specifically, I:

- compared profiles of the temperature of maximum expression for thermally-responsive transcripts
- compared basal expression at optimal temperature for thermally-responsive genes
- compared degree of upregulation to constitutive expression
- compared the critical temperature of transcript upregulation, T~on~, between species for transcripts induced in response to high or low temperatures 
- compared the critical temperature of transcript downregulation, T~off~, between species for transcripts induced in response to high or low temperatures

For these tests, I predicted expression for responsive transcripts.

@@

Probability density function of peak expression for transcripts that differ in expression between *A22* and *Ar*. For this analysis, I used only the transcripts with a significant temperature by species interaction.

@@

The `t.test` fails to account for the many orders of magnitude difference in expression among transcripts, e.g. non-equal variances. This problem is the key issue in the analysis of differential expression `r citep(c("10.1186/1471-2105-11-94", "10.1038/nprot.2013.099"))`. As my goal is simply to determine if expression is typically greater at optimal temperatures (19.5 C) in *Ar* than *A22* for genes that are up-regulated at high temperatures in *A22*, I use a non-parametric Wilcoxon signed rank-test.

@@

Consistent with expectations, there is greater expression at 25C for *Ar* than *A22* transcripts for the set of transcripts that are transcripts that are up-regulated at high temperatures in *A22*. Note that *A22* had the larger library size so if this was due to TPM not correctly accounting for differences in reads between samples, we would expect to see a positive instead of negative value here.

@@

I extracted the **High** transcripts for *A. picea* and examined the log ratio of inducibility against the log ratio of constitutive expression in *A. picea* over *A. carolinensis**. According to the *genetic assimilation hypothesis* there should be a negative relationship between these ratios. That is, transcripts with greater constitutive expression in *A. carolinensis* should have greater inducibility in response to **High** temperatures in *A. picea*.

@@

As with the previous comparison, I also found no evidence for the genetic assimilation hypothesis.

@@

Thermally-responsive genes could also differ in the critical temperature of gene induction. To examine this, I determined the temperature at which each responsive gene had the greatest increase or decrease in expression.

@@

For next analyses, I extracted the list of gene names by response type.

@@

I calculated T~on~ for *High* genes in each species using the observed (rather than predicted) values for each species as by nature of fitting a quadratic function, the maximum change tends to be at the extremes (38.5 or 0) for the predicted values.

@@

This test revealed no differences among species in T~on~ for *High* genes.

I performed the same test comparing T~on~ for *Low* genes.

@@

As with genes turned on at high temperatures, *A. carolinensis* **Intermediate** genes are down-regulated on at higher temperatures than *A. picea* **Intermediate** genes. Towards low temperatures, the mean temperature of downregulation was not different between species.

@@

I combined these gene activation comparisons into a single plot.

@@

Next, I tested if the mean expression level for all responsive genes differed among colonies.

@@

On average, *A. carolinensis* has higher expression than *A. picea*.

@@

```{r ann_responsive, echo=TRUE}
responsive.lms.ann.type <- signif.transcripts[names(responsive.lms)]
# combine responsive.lms.ann with "type" information
cols <- c("Transcript", "A22.type", "Ar.type")
subtab <- sig.response.type[, cols, with = FALSE]
setkey(subtab, "Transcript")
responsive.lms.ann.type <- base::merge(responsive.lms.ann.type, subtab, by = "Transcript", all = TRUE)
str(responsive.lms.ann.type)
write.csv(responsive.lms.ann.type, file = paste(resultsdir, "Ap_responsive_genes_", Sys.Date(), ".csv", sep = ""), row.names = FALSE)
```

@@

There are no heat shock protein-related genes in the responsive transcripts.

@@

I used [topGO](http://www.bioconductor.org/packages/2.12/bioc/html/topGO.html) to perform gene set enrichment analysis (GSEA) seperately for each expression type (bimodal, intermediate, high, low). With the GSEA results, I applied the `selectFDR` function to select transcripts with adjusted P < 0.05. For many of the responsive categories, it appeared that there is considerable redundancy in the enriched GO terms. I used the [GOSemSim](http://www.bioconductor.org/packages/release/bioc/html/GOSemSim.html) package to determine semantic similarity of enriched GO terms `r citep("10.1093/bioinformatics/btq064")` and then performed hierarhichal clustering based on this information distance.

For the GSEA, I first created a gene ID to GO term map file.

@@

Then I read the map file into memory.

@@

Using this gene2GO map file, I performed GSEA for:

@@

@@

As the molecular processes involved in cold and hot temperature tolerance are different, I performed GSEA separately for each response type and compiled results in a single table.

```{r gsea_Ap_high, results = 'hide'}
sig.response.type <- as.data.frame(sig.response.type)
## Genes with *High* expression in both colonies
Ap.high <- sig.response.type[which(sig.response.type$Ar.type == "High" & sig.response.type$A22.type == "High"), "Transcript"]
Ap.geneList.high <- rep(0, length = length(Ap.geneList))
names(Ap.geneList.high) <- names(Ap.geneList)
Ap.geneList.high[(which(names(Ap.geneList.high) %in% Ap.high))] <- 1
# check correct number of values set to 1
table(Ap.geneList.high)
# run GSEA
Ap.high.gsea <- gsea(genelist = Ap.geneList.high, geneID2GO = geneID2GO, plotpath = NA)
```

A single GO term, *`r paste(Ap.high.gsea[, 1:2], collapse = " ")`*, is enriched for **High** genes in both species.

```{r gsea_Ap_low, results = "hide"}
## Genes with *Low* expression in both colonies
Ap.low <- sig.response.type[which(sig.response.type$Ar.type == "Low" & sig.response.type$A22.type == "Low"), "Transcript"]
Ap.geneList.low <- rep(0, length = length(Ap.geneList))
names(Ap.geneList.low) <- names(Ap.geneList)
Ap.geneList.low[(which(names(Ap.geneList.low) %in% Ap.low))] <- 1
# check correct number of values set to 1
table(Ap.geneList.low)
# Run GSEA
Ap.low.gsea <- gsea(genelist = Ap.geneList.low, geneID2GO = geneID2GO)
```

The GO terms, *`r paste(Ap.low.gsea[, 2], collapse = ", ")`*, are enriched for **Low** genes in both species.

```{r gsea_Ap_bim, resuts = 'hide'}
# Genes with *Bimodal* expression in both colonies
Ap.bim <- sig.response.type[which(sig.response.type$A22.type == "Bimodal" & sig.response.type$Ar.type == "Bimodal"), "Transcript"]
# create gene list, setting value to 1 for "bim" transcripts
Ap.geneList.bim <- rep(0, length = length(Ap.geneList))
names(Ap.geneList.bim) <- names(Ap.geneList)
Ap.geneList.bim[(which(names(Ap.geneList.bim) %in% Ap.bim))] <- 1
# check correct number of values set to 1
table(Ap.geneList.bim)
# Run GSEA
Ap.bim.gsea <- gsea(genelist = Ap.geneList.bim, geneID2GO = geneID2GO)
```

The GO terms, *`r paste(Ap.bim.gsea[, 2], collapse = ", ")`*, are enriched for **Bimodal** genes in both species.
